---
title: "Monoculture Bayesian models"
output: html_document
date: "2024-05-03"
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r, label = "git access token for updating, valid until 5.2025", echo = FALSE}
# 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
```

First loading necessary packages - just in case I'm loading JAGS and stan, even though my main goal is to use stan, as it allows a more detailed approach:

```{r, label = "Package loading"}
library(rstan)
library(rjags)
library(R2jags)
library(rmutil)
library(boot)
library(loo)
```

## The data
The data frames are in my R session thanks to the 'Monoculture data.Rmd' file. They are called H_123, B_123,.... The first letter is the species in question, and the '123' represent the 3 years that are shown. The DFs have a simple structure - 'Nt', 'yvalues' which is Nt+1/Nt, and 'year' which represents the year in which Nt+1 was sampled. 

## Our goal
We'll try and estimate $\lambda$ and $\alpha$, both parameters in the Beverton-Holt equation:

\begingroup\Large
\begin{equation*}
\frac{N_{t+1}}{N_{t}} = \frac{\lambda}{1 + \alpha N}
\end{equation*}
\endgroup

We'll approach this in a Bayesian framework, as it allows the incorporation of prior knowledge and an estimation of the uncertainty. 

I'm first using JAGS. This is a simple model and it can probably handle it. The model itself will be written as: 

```{r, label = "JAGS text"}
jags_mono_model <- '
model {
  for(i in 1:N) {
    y[i] ~ dnorm(lambda/(1+alpha*x[i]), residual_sd^-2)
  }
  lambda ~ dunif(0, 5000)
  alpha ~ dunif(0, 4)
  residual_sd ~ dunif(0, 1000)
}
'

```

Now I'll run the model for our most well-behaved species (Rumex), in the hope that the rest will follow:
```{r, label = "Running the model for Rumex"}
model_parameters <- c("lambda", "alpha", "residual_sd")

rumex_model_run <- jags(data = list(N = nrow(R_123),
                                    y = R_123$yvalues,
                                    x = R_123$Nt),
                        parameters.to.save = model_parameters,
                        model.file = textConnection(jags_mono_model), 
                        n.iter = 20000)

rumex_posterior <- as.data.frame(rumex_model_run$BUGSoutput$sims.matrix)
print(rumex_model_run)
```


```{r, label = "Print results", echo = FALSE}
cat("Rumex DIC =", rumex_model_run[["BUGSoutput"]][["DIC"]])

hist(rumex_posterior$alpha)
hist(rumex_posterior$lambda)
hist(rumex_posterior$residual_sd)
hist(rumex_posterior$deviance)

traceplot(rumex_model_run)

rumex_al <- rumex_posterior[ , c("alpha", "lambda")]

alphaq <- quantile(rumex_al$alpha, probs = seq(0, 1, by = 0.1))
install.packages("hexbin")
library(hexbin)
library(ggplot2)

h@count

ggplot(rumex_al, aes(x = alpha, y = lambda)) + geom_hex(bins = 80) + 
  scale_fill_gradient(low="lightblue1", high="darkblue") + 
  theme(aspect.ratio = 1)
```


```{r, label = "Extracting WAIC", echo = FALSE}
samples_normal <- jags.samples(rumex_model_run$model, 
                           c("WAIC","deviance"), 
                           type = "mean", 
                           n.iter = 20000,
                           n.burnin = 1000,
                           n.thin = 1)
samples_normal$p_waic <- samples_normal$WAIC
samples_normal$waic <- samples_normal$deviance + samples_normal$p_waic
tmp_normal <- sapply(samples_normal, sum)
waic_normal <- round(c(waic = tmp_normal[["waic"]], 
                       p_waic = tmp_normal[["p_waic"]]),1)

samples_uni <- jags.samples(rumex_model_run_un$model, 
                           c("WAIC","deviance"), 
                           type = "mean", 
                           n.iter = 20000,
                           n.burnin = 1000,
                           n.thin = 1)
samples_uni$p_waic <- samples_uni$WAIC
samples_uni$waic <- samples_uni$deviance + samples_uni$p_waic
tmp_uni <- sapply(samples_uni, sum)
waic_uni <- round(c(waic = tmp_uni[["waic"]], 
                       p_waic = tmp_uni[["p_waic"]]),1)

cat("Normal distribution WAIC =", as.numeric(waic_normal[1]), "\n")
cat("Uniform distribution WAIC =", as.numeric(waic_uni[1]), "\n")



```